%-------------------------------------------------------------------------------
% Abstract
%-------------------------------------------------------------------------------

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Modern engineering design increasingly demands configurations tailored to specific multi-objective performance profiles, yet discovering these optimal decisions is computationally bottlenecked by the extreme cost of forward physical evaluations. Consequently, inverse design must often operate under severe data sparsity, a regime where traditional deep generative models struggle to resolve the complex, many-to-one relationship between target outcomes and physical decisions. When subjected to highly restricted data budgets, standard probabilistic approaches can suffer from latent space collapse or miscalibration, inadvertently proposing candidate designs across invalid physical voids. To address these limitations, this thesis proposes a novel, geometrically bounded framework for data-driven inverse decision mapping. Rather than relying solely on unconstrained global density estimation, the proposed approach anchors candidate generation to exact local geometric structures and dynamically bounded trust regions, aiming to keep proposed solutions within a mathematically safe physical manifold. Comprehensive empirical evaluations—utilizing a dual-perspective diagnostic suite that measures both objective-space plausibility and decision-space statistical reliability—reveal distinct functional trade-offs between the evaluated architectures. While standard baselines such as the neural networks retain advantages in overall probabilistic capacity, the proposed geometric framework provides a critical alternative for achieving strict micro-precision and physical safety in data-starved regimes. Ultimately, by characterizing the fundamental balance between continuous parametric probability and exact local geometry, this research provides a rigorous foundation to guide the selection of inverse models based on specific engineering tolerances and available computational budgets.