\chapter{Related Work}\label{chap:related}

The transition from forward performance evaluation to target-driven design necessitates a fundamental reversal in how the relationship between decision and objective spaces is exploited. While traditional methodologies focus on the forward mapping of candidate configurations to their respective outcomes, this thesis is situated within a growing body of research that prioritizes the inverse task: identifying the decision-space origins of specified objective-space patterns. To clarify the specific role of the methodology proposed in this work, this chapter organizes the literature according to the operational context of the inversion. We first examine methods that utilize inverse reasoning as an internal search operator within optimization loops; we then discuss the development of independent inverse surrogates designed for interactive, post-hoc querying; we review generative and probabilistic formulations that address the mathematical non-uniqueness of these mappings; and finally, we examine classical geometric and bounded optimization techniques that ensure physical safety in sparse-data regimes.

A first strand of research integrates inverse modeling directly into the optimization loop. Gholamnezhad et al. observe that while early evolutionary multi-objective optimization (EMO) procedures searched in decision space through variation and selection, modern model-based approaches broaden this template via inverse models that map from objective space back to decision space \cite{Gholamnezhad2022}. Cheng et al. established this inverse-in-loop concept through the IM-MOEA framework, which fits Gaussian-process inverse models on non-dominated solutions and applies random grouping to decompose multivariate inverse relations \cite{Cheng2015IMMOEA}. Farias et al. further extended this principle to decomposition-based and constrained settings (IM-MOEA/D and IM-C-MOEA/D), aligning generation with structured subproblems and explicit constraint handling \cite{FariasA21}. For many-objective problems, Gholamnezhad et al. subsequently proposed replacing Gaussian processes with random-forest regression and introduced uniform reference points to map non-dominated objective vectors more evenly back into decision space \cite{Gholamnezhad2022}. These methods demonstrate that inverse reasoning can accelerate exploration, though they remain tied to population dynamics.


A second strand treats inversion as an offline capability trained from archived data. Gholamnezhad et al. argue that this shift changes the focus from improving an optimizer’s internal search to supporting rapid, repeated queries at evaluation time, aligning more closely with exploratory engineering workflows \cite{Gholamnezhad2022}. Because the inverse map is generally non-unique, researchers emphasize that trustworthy use depends on pairing inverse proposals with feasibility screening and forward validation, whether through simulation or a forward surrogate. This approach enables real-time inverse exploration while keeping the design loop grounded through consistency checks.

A complementary line of work addresses non-uniqueness explicitly by learning conditional distributions. Debnath et al. and Ma et al. demonstrate that conditional variational autoencoders and conditional generative adversarial networks can represent one-to-many mappings and sample diverse candidates for the same target specification \cite{Debnath2021JMI, Ma2019AdvMater}. Ardizzone et al. introduced Invertible Neural Networks (INNs) as a mathematically rigorous framework for addressing ill-posed inverse problems through bijective mappings, utilizing latent variables to capture the full posterior distribution of decisions \cite{Ardizzone2019INN}. In the domain of physical design, Ha et al. successfully applied these principles to map prescribed mechanical responses back to manufacturable architectures \cite{Ha2023NatComm}. 

Whang et al. and Zhang and Mikelsons have recently advanced the reliability of these models by utilizing Normalizing Flows to provide explicit uncertainty quantification \cite{Whang2021, Zhang2022IMFlow}. Most recently, industry-leading research has transitioned toward Diffusion Probabilistic Models to overcome stability limitations; for instance, recent frameworks demonstrate that preference-guided diffusion can generate superior Pareto-optimal candidates while maintaining high alignment with multi-objective specifications \cite{PreferenceDiffusion2025}. Furthermore, as noted in recent studies on adjoint-based diffusion, there is a growing move toward physics-informed constraints to ensure that generated decisions remain physically realizable \cite{AdjointDiffusion2026}.

Finally, Dong and Zeng address a different inverse task through Inverse Multi-Objective Optimization (IMOP). Rather than mapping targets to decisions, they focus on inferring the parameters of a decision-making problem from observed Pareto-optimal choices \cite{DongZeng2020ICML, DongZeng2021AAAI}. Their distributionally robust formulation, WRO--IMOP, formalizes the robustness and identifiability concerns that arise when data is limited—a perspective that motivates the safeguards used in this thesis for inverse querying.

Returning to the direct mapping of targets to decisions, while the deep generative models discussed above offer powerful representations for data-rich environments, their reliance on asymptotic statistical guarantees renders them highly vulnerable to generative hallucination in sparse-data regimes. To address extreme data sparsity, optimization literature traditionally pivots toward geometric interpolation and surrogate-assisted bounded search. Techniques utilizing Radial Basis Functions (RBFs) or Kriging frequently employ Trust-Region frameworks to mathematically bound the optimization step, actively preventing the solver from extrapolating into non-physical topological voids \cite{Conn2000TrustRegion, Regis2005Constrained}. Furthermore, spatial partitioning algorithms, most notably Delaunay triangulation and Barycentric coordinates, have been extensively utilized in response surface methodologies to construct safe, deterministic interpolations between known discrete design anchors \cite{Beliakov2006, Jones1998EGO}. However, traditional geometric methods are inherently deterministic and struggle to capture the multi-modal, one-to-many diversity required for true inverse design. Bridging rigid topological safety with diverse generative sampling remains a critical open challenge in the literature.

In summary, the literature reveals a clear dichotomy in inverse design: optimization-centric algorithms provide strict geometric safety but lack generative diversity, while deep probabilistic models (such as CVAEs and INNs) excel at generating diverse candidate sets but fail to respect physical boundaries when training data is sparse. This thesis is situated at the exact intersection of these two paradigms. By adopting the strategy of an independent inverse surrogate, we leverage the baseline generative capabilities of probabilistic models for data-rich scenarios. Simultaneously, to resolve the critical gap in sparse-data reliability, this thesis introduces the Geometrically Bounded Probabilistic Inversion (GBPI) framework. By hybridizing exact Delaunay meshing and Trust-Region optimization with probabilistic Dirichlet sampling, this research proposes a novel synthesis that guarantees both physical safety and multi-modal candidate generation in highly constrained environments.