%-------------------------------------------------------------------------------
% Discussion
%-------------------------------------------------------------------------------

\chapter{Discussion}\label{chap:discussion}

This chapter synthesizes the empirical findings from Chapter~\ref{chap:experiments}, relates them to the evaluation criteria introduced in Chapter~\ref{chap:methodology}, and discusses the broader implications, limitations, and directions for future research in autonomous inverse design.

\section{Interpreting the Findings and Implications}\label{sec:discussion_interpretation}

The synthesis of empirical results from the $f_5$ experiments reveals that architectural selection for inverse design is not merely a matter of hyperparameter tuning, but a fundamental choice between competing mathematical paradigms. By analyzing objective-space precision (ECDF), probabilistic calibration (MACE, CRPS), and geometric scaling across dimensions, this study demonstrates a critical vulnerability in modern deep generative models: they are structurally incompatible with extreme data sparsity.

A primary finding of this work is the stark divergence between parametric probability estimation and exact geometric bounding. When restricted to a highly constrained data budget ($N=100$), the deep learning baselines face a severe trade-off between expressive capacity and structural collapse. The CVAE models exhibited severe pathology; forced to compress sparse data into a continuous Gaussian prior, their latent spaces collapsed. This resulted in a persistent S-curve signature in their PIT calibration plots (indicating severe overconfidence) and a fundamental inability to achieve micro-precision in the objective space. 

Conversely, while the MDN and INN architectures achieved better probabilistic calibration (MACE) and continuous accuracy (CRPS) in the 2D decision space, the 5D stress test exposed their inherent reliance on dense data. Subjected to the curse of dimensionality, their continuous probability manifolds "smeared" across vast topological voids, yielding blurry, macroscopic approximations rather than precise engineering solutions.

In direct contrast to these parametric vulnerabilities, the proposed Geometrically Bounded Probabilistic Inversion (GBPI) framework proved that replacing global probability distributions with localized geometric constraints ensures robust, high-fidelity target attainment. By relying on exact Delaunay meshing and Trust-Region boundaries, GBPI maintained its micro-precision (registering exact target matches below the $1\mu$ scale) entirely independent of the surrounding dimensional voids. 

Based on the statistical and geometric signatures observed throughout the experiments, the evaluated frameworks can be categorized into three distinct functional profiles:

\begin{itemize}
    \item \textbf{The Geometric Standard (GBPI):} The proposed framework emerges as the most robust engine for high-stakes, data-sparse inverse problems. By refusing to map continuous probabilities across empty topological voids, it systematically prevents physical hallucination. Its mathematically guaranteed physical safety and dimensional resilience make it the optimal choice for autonomous propose--verify loops where structural integrity and micro-precision are paramount.
    \item \textbf{The Macroscopic Approximators (MDN \& INN):} These models serve as capable tools for low-dimensional, macroscopic exploration. The INN tracks the ideal calibration diagonal well in 2D, making it statistically honest about its own uncertainty. However, their reliance on global density estimation makes them "blurry" at the micro-scale and highly vulnerable to the curse of dimensionality, rendering them unsafe for strict tolerance requirements in complex design spaces.
    \item \textbf{The Data-Starved Priors (CVAE):} While standard Variational Autoencoders are highly effective in dense data domains (e.g., image generation), this study proves they are fundamentally ill-suited for strict sparse-data engineering tasks. Their inherent architectural pressure to match a simplified prior distribution causes latent collapse under sparsity, resulting in massive attainment lag and unreliable uncertainty estimates.
\end{itemize}

These results conclusively suggest that the choice of inverse model must be dictated by data availability and tolerance requirements. If the goal is rapid, automated attainment within microscopic physical tolerances using highly restricted data, geometric boundaries are not just advantageous—they are mathematically necessary. Ultimately, the achievement of this diagnostic suite is the ability to separate models that merely output "probable" guesses from a framework that guarantees physically bounded, mathematically exact solutions.

\begin{table}[htbp]
\centering
\renewcommand{\arraystretch}{1.4} 
\setlength{\tabcolsep}{6pt}        
\caption[Decision Matrix: Model Performance Summary]{Decision Matrix: Summary of architectural capabilities under extreme data sparsity ($N=100$). GBPI explicitly trades global probabilistic density estimation for absolute geometric safety and micro-precision.}
\label{tab:decision_matrix}
\begin{tabularx}{\textwidth}{
    >{\raggedright\arraybackslash}l 
    >{\centering\arraybackslash}m{2.5cm}
    >{\centering\arraybackslash}m{2.5cm}
    >{\centering\arraybackslash}m{2.5cm}
    >{\raggedright\arraybackslash}X}
\toprule
\textbf{Model Family} 
& \textbf{\shortstack{Micro- \\ Precision}} 
& \textbf{\shortstack{Dimensional \\ Scalability}} 
& \textbf{\shortstack{Physical \\ Safety}} 
& \textbf{Primary Characteristic} \\
\midrule
\textbf{GBPI (Proposed)} & Dominant & Robust & Guaranteed & Geometric Exactness \\
\textbf{INN}             & Marginal & Degraded & Unbounded & Statistical Honesty \\
\textbf{MDN}             & Marginal & Collapsed & Unbounded & Mass Concentration \\
\textbf{CVAE}            & Failed   & Collapsed & Unbounded & Latent Compression \\
\bottomrule
\end{tabularx}
\end{table}


%-------------------------------------------------------------------------------
\section{Limitations and Assumptions}\label{sec:discussion_limits}
%-------------------------------------------------------------------------------

The evaluation framework defined in Section~\ref{sec:evaluation} provides a rigorous statistical and geometric basis for assessing inverse models under severe data sparsity. However, interpreting these results in a physical materials-discovery or engineering setting requires acknowledging foundational assumptions and limitations that reflect the gap between computational attainment and experimental reality.

\begin{itemize}
  \item \textbf{Fidelity of the Forward Evaluator ($\mathcal{E}$) and Surrogate Reliance:} 
  The propose--verify contract assumes that the evaluator $\mathcal{E}$ functions as a reliable ground-truth proxy. In this study, due to the strict computational budget restricting the dataset to $N=100$ exact physical evaluations, an internal Radial Basis Function (RBF) surrogate is employed as $\mathcal{E}$ to score the generated candidates. A primary limitation is that computational attainment certifies validity only with respect to this RBF surface, not absolute physical laws. If the RBF surrogate inaccurately interpolates the topological voids between the sparse anchors, an inverse design proposed by GBPI or the neural baselines may appear highly precise computationally but still fail in experimental synthesis. In this case, the reported ECDF attainment reflects internal mathematical alignment with the surrogate rather than absolute physical certainty.

  \item \textbf{Completeness of the Decision Space Representation ($\mathcal{X}$):} 
  The methodology inherently assumes that the decision vector $\vect{x}$ provides a sufficient and complete description of the physical system to determine the properties in $\mathcal{Y}$. A severe limitation arises when combining a potentially incomplete physical feature set with the extreme sparsity of the training archive. Real-world performance often depends on unobserved physical factors—such as microstructure, thermal history, or explicit synthesis pathways—that are absent from $\vect{x}$. In dense datasets, generative models can sometimes average out these hidden variables. However, under extreme data restrictions, models cannot statistically compensate for missing physics. Consequently, an inverse model may propose a candidate that is mathematically coherent within the normalized coordinate system, but physically underdetermined or structurally unstable in reality.

  \item \textbf{Extrapolation Beyond the Training Manifold:} 
  A fundamental limitation in inverse modeling is the degradation of reliability when querying targets that lie entirely outside the known domain of the training data. Because each algorithm in this study treats extrapolation differently, establishing a unified performance metric in these unmapped regions is complex. Deep generative models rely on unbounded continuous probabilities, often resulting in unconstrained geometric hallucination when pushed past the training data. Conversely, the proposed GBPI framework mitigates this by transitioning from exact interpolation to a bounded Trust-Region optimization (Pathway B). While this boundary enforces physical safety, it means the extrapolation behavior is highly dependent on the algorithm's specific architectural fail-safes rather than a shared probabilistic baseline.

  \item \textbf{Dimensional Scalability of Exact Geometric Meshing:} 
  While Chapter~\ref{chap:experiments} demonstrated that the GBPI framework possesses extraordinary resilience when scaling the decision space ($D$) from 2D to 5D, its reliance on exact computational geometry introduces a hard scalability limit regarding the objective space ($M$). Specifically, the Delaunay triangulation—which drives the target-enclosing mechanism of Pathway A—is computed entirely in the $M$-dimensional objective space. The time complexity of this meshing operation scales worst-case as $\mathcal{O}(N^{\lceil M/2 \rceil})$. Furthermore, if the number of simultaneous objectives becomes excessively large (e.g., $M > 10$), the mathematical concept of enclosing local simplices degrades, as nearly all data points migrate to the boundaries of the Convex Hull. Therefore, a fundamental limitation of the proposed GBPI framework is that it is strictly optimized for physical engineering problems with a manageable number of performance criteria ($M \le 5$) and low-to-medium decision parameters ($D \le 10$). It is not designed for ultra-high-dimensional target spaces (like raw image or text generation) where deep latent-variable models remain computationally necessary.
  
\end{itemize}


%-------------------------------------------------------------------------------
\section{Future Work}\label{sec:future_work}
%-------------------------------------------------------------------------------

Building on the findings from Chapters~\ref{chap:methodology} and \ref{chap:experiments}, several concrete extensions could address the limitations of geometric scaling and further improve the reliability of autonomous inverse querying.

\begin{itemize}
  \item \textbf{Approximate Geometry for Many-Objective Scaling:} 
  The primary mathematical bottleneck of the GBPI framework is the $\mathcal{O}(N^{\lceil M/2 \rceil})$ time complexity of exact Delaunay triangulation in the objective space ($Y$-space). Future work should investigate integrating approximate geometric structures, such as Alpha-shapes ($\alpha$-complexes), spatial hashing, or localized k-Nearest Neighbor (k-NN) sub-meshes. By constructing geometric boundaries only locally around the target query rather than computing the global Delaunay complex, GBPI could scale efficiently to many-objective engineering problems ($M > 5$) without losing its guarantee of physical safety mapping back to the decision space.

  \item \textbf{Topological Data Analysis (TDA) for Manifold Bounding:} 
  Currently, the extrapolative bounds of GBPI rely on localized Euclidean distance thresholds within the Trust-Region optimizer (Pathway B). However, in highly complex decision spaces, a simple Euclidean radius might inadvertently bridge severe non-linearities, leading the optimizer to cross through invalid physical voids.  Future implementations should advance the framework by integrating Topological Data Analysis (TDA)—specifically replacing the Euclidean threshold with a Vietoris-Rips structural graph check. By computing the persistent homology of the sparse training archive, the algorithm could explicitly detect non-linear manifolds and topological holes, ensuring that candidate generation strictly conforms to the true topological skeleton of the physical problem rather than assuming isotropic Euclidean safety.
  
  \item \textbf{Surrogate-Aware Geometric Bounding:} 
  Because the propose--verify loop in sparse regimes relies heavily on an internal surrogate (e.g., the RBF), future iterations of GBPI should explicitly couple the Trust Region radius to the surrogate's local epistemic uncertainty.  By upgrading the deterministic RBF to a probabilistic surrogate (such as a Gaussian Process), the optimization boundary could dynamically contract in regions where the predictive variance is high, and expand in regions where the surrogate is highly confident. This would create a "physics-aware" geometry that actively protects the inverse solver from exploiting artificial optima or hallucinations hidden in the surrogate landscape.
\end{itemize}

%-------------------------------------------------------------------------------
\section{Conclusion}\label{sec:conclusion}
%-------------------------------------------------------------------------------

This thesis established a robust framework for autonomous inverse design, specifically addressing the critical bottleneck of extreme data sparsity in engineering applications. By investigating both 2D and 5D decision spaces under a strict $N=100$ data budget, this study demonstrated that the fundamental relationship between physical decisions and performance outcomes cannot be safely navigated using traditional deep generative models alone. 

The central contribution of this work is the development and validation of the Geometrically Bounded Probabilistic Inversion (GBPI) framework. This research proved that in data-starved environments, architectural selection is not merely a matter of capacity, but a choice between competing paradigms: continuous parametric probability versus exact local geometry. 

Our exhaustive empirical evaluation revealed that while deep learning baselines (CVAE, MDN, INN) can exhibit statistical reliability in low-dimensional spaces, they suffer from catastrophic structural degradation when subjected to the curse of dimensionality. Forced to parameterize high-dimensional probability distributions with insufficient spatial data, these continuous models succumbed to latent space collapse, severe overconfidence, and geometric hallucination. They essentially "smeared" probability mass across vast topological voids, resulting in blurry, macroscopic approximations that consistently failed to hit targets at rigorous micro-precision tolerances.

In stark contrast, the proposed GBPI framework completely bypassed the requirement for dense global probability mapping. By anchoring candidate generation strictly to exact Delaunay meshing in the objective space and dynamically bounded Trust-Region optimization in the decision space, GBPI successfully locked onto verified physical realities. The objective-space diagnostics definitively showed that GBPI dominated the micro-precision regime, registering exact target matches at the nanometer scale while systematically preventing the generation of physically impossible candidates.

Ultimately, this thesis concludes that standard objective-space alignment and probabilistic calibration are insufficient metrics when the underlying physical manifold is sparse and disjointed. In high-stakes, severely constrained engineering workflows, relying on unconstrained global probabilities is fundamentally unsafe. By proving that exact geometric boundaries are vastly superior to scaled-up parametric capacities, this research provides a definitive, reliable engine for autonomous materials discovery and target-driven design.